Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary

### 2024-05-23 revision/review_response.qmd

---
format: 
  docx:
    reference-doc: response_styles.docx
bibliography: ../refs.bib
csl: ../estuaries-and-coasts.csl

execute: 
  echo: false
  warning: false
  message: false
---

```{r}
#| eval: false
spelling::spell_check_files(here::here('revision/review_response.qmd'))
```

*We sincerely thank the associate editor and all reviewers for providing useful comments on our manuscript. We have made every effort to address these concerns. In particular, we have provided a more thorough assessment of seagrass response to climate-related stressors using more flexible models (GAMs) and including additional predictors (light attenuation). Please see the responses below (line numbers in responses refer to the revised manuscript).*

@@

2\. Justification for period of decline: I struggle with the assertion that 2016 represents the beginning of the dramatic seagrass loss across Tampa Bay, as stated on Line 183. This is especially problematic when observing seagrass coverage in Figure 2a, which represents total species coverage. Here, coverage is relatively stable 1988 to 2012, then increases 2014-16, with levels dropping after this. The extent of these patterns differs per bay segment – in OTB, there is a dramatic increase with some decrease, although only the last year (2022) is below the baseline values 1988 to 2012. For other segments, there is a decline but coverage post 2016 is either similar to the baseline or higher. What are the drivers of those large increases in more recent years, and are declines relevant or is seagrass coverage simply returning to baselines levels that were evident 1988-2012? Please comment in the text. Further, the decline in cover is more evident when observing species specific data, particularly for H. wrightii in OTB and MTB, but not other segments. Should the analyses focus on the dominant species driving these patterns? Please revise text and acknowledge these aspects, perhaps reanalyse the data or interpretation if necessary, to present a more convincing justification.

*Response: The above interpretation is correct in that the patterns of increase/decrease vary by bay segment and dataset (coverage or frequency occurrence). This suggests that multiple factors are influencing these changes and that they potentially vary by location. Our development of more robust models (GAMs) that attempted to better describe these patterns is an explicit acknowledgment of these differences, which were not adequately characterized with the earlier models. Please see our description below of these updated models.*

*Additional text was also added to clarify these nuanced changes and the focus on total species occurrence.*

*Line 197: “As such, trajectories of recovery and decline have varied by bay segment in magnitude and timing of the change, although a consistent decline baywide has been observed since 2016. Change by species has been most notable for Halodule wrightii (shoal grass) in OTB and HB, whereas some losses have been observed for Thalassia testudinum (turtle grass) in LTB, likely related to dominance of each species across the salinity gradient and proximity to hydrologic inputs (Lewis et al. 1985).”*

*Line 372: “Additionally, total seagrass frequency occurrence was used as a response variable as an aggregate measure of community change to potential stressors given the likelihood that species-specific changes may be more difficult to model and that the majority of seagrass loss in recent years was dominated by H. wrightii in the upper bay segments (Figure 2b).”*

@@

Line 205: Define frequency occurrence.

*Response: Text revised as “...cover-abundance, frequency occurrence (number of sample points with seagrass divided by total points on a transect, as in Sherwood et al. 2017), and condition, ...”*

Line 207: My understanding is that both transects and areal maps were done once a year. Clarify here why transect data provided greater temporal resolution.

*Response: As noted on line 206, the areal maps are produced every other year, whereas the transect data are collected every year. The text was revised as follows: “Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year. The transect data are collected each year, allowing inter-annual comparison...”*

@@

Lines 303-305 "we focus on water temperature and salinity given that other dominant forcing factors, i.e., light availability, have been relatively stable in recent years." Here a citation is given, however these mentioned conditions appear to be stable for areas except OTB, which is also the area that saw the greatest declines in seagrass. So again, why leave light availability out of the models?

*Response: Please see our above responses.*

Lines 225-226 "Most samples are collected from mid-morning to early afternoon". That is a major issue with using temperature data that is a snapshot and not continuous. Temperatures can vary greatly between morning to afternoon. Were any attempts made to normalize temp data to time of day? I don't know if it would be possible given the datasets, but something to consider.

*Response: This is certainly an issue worth considering using the temperature data from the monthly discrete samples and was the motivation for including a statement on the need for continuous monitoring data in the original draft (lines 679 - 681). Applying a correction or normalization for time of day would be challenging and we would not be confident in the results to accurately describe maximum or minimum temperature values for a given day. Moreover, we feel the trends in the temperature summaries (e.g., increase in number of days above a threshold over time) would likely be similar. As such, we have not attempted this for the current analysis. However, ongoing work in Old Tampa Bay for a separate project is attempting to describe short-term diurnal variation in temperature with continuous data loggers. We suspect this information will provide further insight into acute temperature stressors that could affect seagrass. We have added text describing this general need and a statement with a short description of this work, although these loggers are in very shallow areas that likely show different temperature characteristics than the discrete samples at deeper depths.*

*Line 582: Without more continuous, diel observations of these metrics over the period of record, we were heavily reliant on these model outputs to determine relevant thresholds for Tampa Bay seagrass. This further highlights a long-standing data gap and need for the estuary."*

*Line 681: “Ongoing work in OTB using continuous data loggers in shallow areas where seagrass has been gained or lost will provide insights into short-term diurnal changes as potential acute temperature stress (see https://tbep-tech.github.io/otb-temp/tempeval).”*

@@

Lines 313-314 "Shoal grass is tolerant of a wide range of salinity..." Shoal grass accounted for the largest seagrass declines, according to Figure 2b, so why would decreases in salinity negatively impact this species? Wouldn't a decrease of 0.04-0.06 per year be well within the tolerance range for this species?

*Response: Our previous statement on lines 571-574 suggests this may be a plausible explanation for the weak association of seagrass change with the stressors. The statement was further revised to specifically mention that shoal grass accounted for the largest seagrass decline: “This is especially true for H. wrightii that had the greatest changes over the period of record and is tolerant of a wide range of salinity.”*

@@

Lines 350-352: Did you experiment with different lagged responses?

*Response: Yes, preliminary analyses evaluated lagged associations between the salinity and temperature stressor metrics and seagrass change. Specifically, we evaluated the number of days in 30 day periods from one month to one year when temperature was above or salinity was below a threshold in relation to seagrass change (see code here, https://github.com/tbep-tech/temp-manu/blob/9044bb07cd81532a7fc83c02dc035f7d8ff16b2c/R/dat_proc.R#L865, lines 865 to 925 in the file). No compelling differences were observed for any of the lags compared to the entire year, which was the justification for using the latter in the current analysis. Additional text was added to this effect:*

*Line 360: “Preliminary analyses evaluated different lagged associations between the stressor metrics and seagrass change, although initial results suggested no additional insight could be gained using lagged assessments compared to the transect year summaries. As such, the stressor metrics...”*

@@

Line 384: what was the justification for removing September?

*Response: The assessment of inter-annual changes in precipitation during the rainy season showed a weak increase over time, whereas removing the last month (September) showed a more statistically powerful increase over time. The removal of September was meant to show that the trend was driven by an increase in the first three months of the rainy season (Jun - Aug). Text was added to line 413: “...suggesting precipitation increases were driven by the earlier months (June - August).” Also note that our previous estimates of rainfall total were based on an incorrect sum across overlapping areas. We have fixed the issue, although the results and trends remain the same.*

@@

Before acceptance I recommend that several points should be addressed to improved the manuscript's clarity and usefulness.

*Response: Thank you for providing useful feedback on our manuscript. Please see our detailed responses below.*

First, although light availability is one of the most principal water quality factors affecting seagrass performance, stability and resilience, data is not presented here to characterize the light environment either spatially or temporally, other than to suggest it has been relatively stable and consistent across the system. However, on L. 120 it is highlighted that "... upper bay segments where marginal light environments exist".

*Response: Line 125 was revised to describe that the upper bay segments are more shallow and receive a majority of hydrologic inflow. The previous statement was inaccurate. Also, please see the addition of Figure 2c that shows the long-term improvement in the light environment over time. Of particular note is that all bay segments are currently below the threshold supportive for seagrass growth.*

### 2024-05-23 cover-letter.qmd
ready to resubmit

Enclosed please find our manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.

The long-term recovery of Tampa Bay is a nationally recognized success story that demonstrates how the USEPA National Estuary Program is a valuable model for collaborative management.  Through large-scale reductions in external nutrient loading, seagrasses have recovered in the bay, reaching an all-time high in 2016.  Since that time, seagrasses have declined by nearly 1/3 and questions remain regarding the role of climate-related stressors in contributing to this decline.  We leverage multiple data sources to describe changing water temperature and salinity in the bay and relate these trends to the recent seagrass loss from 2016 to present. No such comprehensive assessment has previously been conducted for Tampa Bay and our results provide context for how climate stressors can confound existing management models, particularly related to nutrient management.

This work has not been published elsewhere, nor is it under consideration for publication anywhere else. Each author has contributed substantially to this work. All authors have contributed to the writing and review of the original draft. Analysis was conducted by the first author, with interpretation of results supported by all other co-authors. All authors have approved the final draft and the submitted product was prepared according to instructions for authors provided on the journal's website.

We greatly appreciate the opportunity to publish this work in Estuaries and Coasts and are confident readers will consider it a valuable contribution. Please feel free to contact us directly should additional information be needed.

>>>>

Enclosed please find our revised manuscript, titled \`Hot and fresh: evidence of climate-related suboptimal water conditions for seagrass in a large Gulf coast estuary', to be considered as an original article in Estuaries and Coasts.

We sincerely appreciate the constructive comments provided by the editors and reviewers and have made every attempt to address these comments in our revision. The largest change to the manuscript was a reanalysis of the potential seagrass response to climate stressors, which included more flexible Generalized Additive Models and inclusion of additional predictors (i.e., time as a continuous variable, light attenuation).  Overall, our conclusions remain the same that climate stressors are potentially driving recent seagrass changes, although our more robust modeling approach provides greater weight of evidence for this conclusion.  Note that we have also made substantial revisions in response to comments from the co-editor in chief regarding statistical descriptions of results to focus more on effect sizes and magnitudes of change.

Again, we greatly appreciate the opportunity to publish this work in Estuaries and Coasts and are confident readers will consider it a valuable contribution. Please feel free to contact us directly should additional information be needed.

### 2024-05-23 manu-draft.qmd
ready to resubmit

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increased temperature and altered precipitation patterns. These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hammer2018; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The environmental stress on seagrasses from acute warming related to climate change was compounded by stress from poor light environments in shallow waters, demonstrating a concerning synergy of stressors most likely to affect Tampa Bay. @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*). Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed *Z. marina* [@hensel2023; also see @bartenfelder2022]. Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@hall2016; @lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers that contribute to a majority of freshwater inflow to the bay. The climate of Tampa Bay is subtropical with warm humid summers and cool less humid winters [@garcia2023]. Unique to Tampa Bay and the entire central Florida peninsula is that this region is within a transition zone from more temperate weather to the north, similar to the rest of the southeastern United States, and more tropical weather to the south [@morrison2006]. El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@schmidt2002; @greening2006; @morrison2006]. The watershed is heavily developed and includes over 3 million people [@tbep0923] with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].

@@

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of what was present in the 1950s. Though Tampa Bay was far from pristine at the time, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since then, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). Light environments have improved (Figure [-@fig-seagrasschg]c) through reduction of external nitrogen loadings by about 2/3 by the 1980s from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].

@@

Two primary sources of data have been used to track seagrass change in Tampa Bay (@tbl-dattab). The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created from aerial images collected specifically to map seagrass and are acquired during a flight window from December and February. The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment. The maps provide a spatial estimate of seagrass cover at the landscape scale and is irrespective of species. Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on cover-abundance, frequency occurrence [number of sample points with seagrass divided by total points on a transect, as in @sherwood2017], and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year. The transect data are collected each year, allowing inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used below for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Separate models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables. Separate smoothed terms for each predictor and interactions with year for each predictor for each bay segment were included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs. The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows.

@@

The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3). The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively. Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (4 to 72 days per year from 1975 to 2022).

@@

The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment. The FIM model (`r sgmodtxt$fimmod`), Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased. The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity. Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).

@@

Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; @nickerson2023; @shiir]. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used). @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, although they rightfully acknowledge the sensitivity of their results to conditions at the start and end of the time series. Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain. Our results showing increases in temperature and decreases in salinity in the fall, winter period provided a finer-scale comparison, where trends were most notable for OTB and northern stations of HB (@fig-kendall), likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels [@weisberg2006]. Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico. The upper limit of our warming estimate for Tampa Bay is comparable. Notably, @carlson2018 suggest a link between historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are likely suboptimal but within the ecological niche for seagrasses. This is especially true for *H. wrightii* that had the greatest changes over the period of record and is tolerant of a wide range of salinity. However, this does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Our thresholds may also be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors. Regardless, our models suggested that temperature and salinity are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect.

@@

Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay. Management actions that have historically been effective may not be able to maintain ecosystem resilience to climatic change. For example, nitrogen load reductions have been effective at restoring seagrass in Tampa Bay [@greening2006; @greening2014]. As Tampa Bay becomes hotter and fresher, current nutrient loads may no longer be effectively assimilated and algal and seagrass ecology dynamics may shift. Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrass in Tampa Bay. These considerations are especially critical for upper parts of Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends appear most pronounced. Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, given the challenges of restoring these long-lived foundation species once lost, ongoing development in the watershed, and the current climate trajectory.

>>>>

Climate change has complicated the understanding of ecosystem response to conventional stressors, presenting new challenges and expectations for how ecological resources will respond to management actions [@statham2012a; @sherwood2013]. In addition to sea-level rise, the most anticipated effects of climate change in coastal environments are increased temperature and altered precipitation patterns. These changes will profoundly alter physicochemical habitats, creating suboptimal or uninhabitable conditions for many species [@madeira2012; @lefcheck2017; @hammer2018; @hensel2023]. @lefcheck2017 evaluated interactive effects of water clarity and rising temperatures on seagrasses with over 30 years of data in Chesapeake Bay. The environmental stress on seagrasses from acute warming related to climate change was compounded by stress from poor light environments in shallow waters, demonstrating a concerning synergy of stressors most likely to also affect Tampa Bay. @moreno-marin2018 produced similar results using a multifactorial experiment that considered temperature, light, and nutrient (nitrogen) availability for North Sea eelgrass (*Zostera marina*). Species shifts are also expected to occur as changing physical conditions decrease the competitive advantages of historically abundant species. In Chesapeake Bay, the abundance of widgeongrass (*Ruppia maritima*) has responded positively to nutrient reduction and is replacing the formerly dominant but now heat-stressed *Z. marina* [@hensel2023; also see @bartenfelder2022]. Changing frequency and severity of precipitation patterns may further alter salinity regimes and with it the distribution and abundance of seagrasses throughout the estuary [@rasheed2011; @webster2021]. These changes may produce hypersaline conditions under periods of prolonged drought or more freshwater conditions with increased storm events. Seagrass species may respond differently under individual or multiple stressors and each watershed will respond differently to climate change [@hall2016; @lefcheck2017; @zhang2023], suggesting place-based empirical assessments will be needed to properly inform management decisions.

@@

Tampa Bay is the largest open-water estuary in Florida covering 400 mi$^2$ (1,036 km$^2$) and the second largest in the Gulf of Mexico. The watershed covers an additional 2,200 mi$^2$ (5,872 km$^2$) with the Hillsborough, Alafia, Manatee, and Little Manatee rivers contributing the majority of freshwater inflow to the bay. The climate of Tampa Bay is subtropical with warm humid summers and cool, less humid winters [@garcia2023]. Unique to Tampa Bay and the entire central Florida peninsula is that this region is within a transition zone from more temperate weather to the north, similar to the rest of the southeastern United States, and more tropical weather to the south [@morrison2006]. El Niño weather events have also been associated with prolonged periods of heavy rainfall contributing to seagrass reductions through increased stormwater nutrient loads [@schmidt2002; @greening2006; @morrison2006]. The watershed is heavily developed and includes over 3 million people [@tbep0923] with 42% of the land as urban or suburban contributing substantial inputs of wastewater and stormwater runoff that can stress bay resources [@beck2023]. The geology of the watershed is rich in phosphates and mining activities have greatly altered the landscape, with notable spills and releases of wastewater that have affected water quality and biological resources [@garrett2011; @beck2022a].

@@

The long-term recovery of seagrass habitats in Tampa Bay since the 1980s is a nationally recognized success story that demonstrates application of a successful management paradigm through the EPA-administered National Estuary Program [@greening2006; @greening2014; @sherwood2017]. From 1988 to 2016, seagrasses increased 79% to 41,655 acres (16,857 ha), surpassing the regional management goal of restoring coverage to 95% of what was present in the 1950s. Though Tampa Bay was far from pristine at the time, aerial imagery was sufficient to estimate a relatively unimpacted condition for seagrass coverage. Since then, the greatest areal coverage expansions were observed in the upper bay segments (OTB, HB, and MTB; Figure [-@fig-seagrasschg]a). Light environments have improved (Figure [-@fig-seagrasschg]c) through a 2/3 reduction of external nitrogen loadings from a peak 1970s estimate of 8.9 x 10$^6$ kg/year, largely from advanced wastewater treatment upgrades and in part from the cumulative effects of habitat restoration and additional stormwater control projects implemented in the watershed [@greening2014; @beck2019].

@@

Two primary sources of data have been used to track seagrass change in Tampa Bay (@tbl-dattab). The Southwest Florida Water Management District (SWFWMD) has estimated areal coverage of seagrasses approximately biennially since the late 1980s (Figure [-@fig-seagrasschg]a, available at <https://data-swfwmd.opendata.arcgis.com/>). These maps are created from aerial images collected specifically to map seagrass and are acquired during a flight window from December and February. The maps are created by photointerpretation of image signatures coupled with a robust field verification and accuracy assessment. The maps provide a spatial estimate of seagrass cover at the landscape scale, irrespective of species. Complementary to the SWFWMD seagrass mapping program is the Tampa Bay Interagency Seagrass Monitoring Program (Figure [-@fig-seagrasschg]b, <https://tampabay.wateratlas.usf.edu/seagrass-monitoring/>). Annual transect surveys have been conducted since 1998 at 62 fixed locations in Tampa Bay, many of which were chosen to target seagrass beds of interest [@tbep0816; @sherwood2017]. This dataset provides species information on cover-abundance, frequency occurrence [number of sample points with seagrass divided by total points on a transect, as in @sherwood2017], and condition, collected at fixed meter marks along a transect extending from the shoreline to the deepwater edge of the seagrass bed. Although the areal maps provide the standard for assessment of restoration goals, the maps are produced every other year. The transect data are collected each year, allowing inter-annual comparison at greater temporal resolution, particularly for the recent period of interest when seagrasses have declined. As such, the transect data were used for comparison with temperature and salinity changes for the major bay segments. Additional sources of seagrass data are described in the next section.

@@

Two other models were constructed to provide an additional weight-of-evidence for the FIM and PDEM temperature and salinity datasets relative to seagrass change. These models used direct measurements of bottom salinity and temperature as independent variables because the stressor metrics could not be calculated using the sampling designs from these monitoring programs (i.e., each sample was a distinct location). GAMs were used for the FIM data to evaluate annual average seagrass percent cover using year, temperature, and salinity as independent variables. Separate smoothed terms for each predictor and interactions with year for each predictor for each bay segment were included as above. GAMs for the PDEM data were constructed similarly, except only Old Tampa Bay was evaluated due to spatial limitations of the data. Neither the FIM nor PDEM models used light attenuation as a predictor variable given that a large percentage of Secchi observations were measured on the bottom (`r onb$fim` and `r onb$pinco`, respectively), providing further support that light environments have not been limiting for seagrasses. Model input data were further subset to include only months from July to October to describe seagrasses during the growing season and to reduce potential seasonal effects. Lastly, all data were averaged annually for the monthly subsets for comparability of sample size (i.e., power) with the EPC models. For the PDEM models, presence/absence was converted to frequency occurrence as the number of sites with seagrass in a year divided by the total number of sites and all data were subset to less than 2 meters to better characterize locations where seagrass occurs. The GAM equations for the FIM (@eq-fimmod) and PDEM (@eq-pincomod) datasets, respectively, were as follows:

@@

The number of days when both temperature was above and salinity was below a threshold also varied by bay segment (Table S3). The number of days when temperature was above 29 $^\circ$C and salinity was below 25 ppt had the largest slopes of 1.4, 1.1, and 0.7 days per year for OTB, HB, and MTB, respectively. Likewise, the average number of days when both temperature was above 29 $^\circ$C and salinity was below 25 ppt from the beginning to the end of the period showed the greatest increase for OTB of 68 days (Table S5; 4 to 72 days per year from 1975 to 2022).

@@

The GAMs for the FIM and PDEM datasets had notable associations with temperature and salinity stress, where an increase in temperature and a decrease in salinity was often associated with a decline in seagrass depending on bay segment. The FIM model (`r sgmodtxt$fimmod`, Figure [-@fig-sgmod2]a, Table S9) showed a slight decrease in seagrass with increasing temperature for HB and MTB, with a similar change as salinity decreased. The PDEM model (`r sgmodtxt$pincomod`, Figure [-@fig-sgmod2]b, Table S10) that evaluated only OTB showed a distinct decrease with seagrass as temperature increased, but no association with salinity. Some of the interaction terms of the predictors with year suggested the relationships with seagrass may have changed over time, but the results were not consistent between bay segments or models (Tables S7 to S10).

@@

Similar regional, long-term changes in coastal waters and estuaries have been observed by others [@carlson2018; @nickerson2023; @shiir]. @nickerson2023 evaluated sea surface trends at a larger spatial scale for Tampa Bay, the West Florida Continental Shelf, and the adjacent Gulf of Mexico. Temperature trends were similar to those herein for Tampa Bay (the EPC dataset was also used). @nickerson2023 also noted that temperature increases in Tampa Bay were most pronounced in the winter, although they rightfully acknowledge the sensitivity of their results to conditions at the start and end of the time series. Our assessment evaluated non-parametric trends (i.e., Kendall tests, less sensitive to outliers) at individual EPC stations and bay segments, whereas @nickerson2023 evaluated the EPC data as an average for the entire bay for consistency of comparison to their larger spatial area and model domain. Our results showing increases in temperature and decreases in salinity in the fall, winter period provided a finer-scale comparison, where trends were most notable for OTB and northern stations of HB (@fig-kendall) – likely related to hydrodynamic characteristics of these segments relative to MTB and LTB that flush more regularly with the Gulf of Mexico. These upper bay segments are more affected by hydrologic inflows (HB), lack of circulation (OTB), or thermal stress related to more rapid warming with shallower depths. The significant reduction in salinity for LTB is also of note, perhaps related to gravitational circulation patterns that export lower salinity water from upstream in the main shipping channels [@weisberg2006]. Additionally, @shiir provided a recent assessment of a 2023 heatwave in south Florida, supported by a 20-year trend assessment that suggested estuaries were warming at nearly double the rate of the Gulf of Mexico. The upper limit of our warming estimate for Tampa Bay is comparable. Notably, @carlson2018 suggest a link between historical seagrass losses in Florida Bay and rapid warming in shallow areas with low surface reflectance.

@@

An additional caveat of our models was the use of "thresholds" to define potential stressor metrics for temperature and salinity on an annual time scale. Our choice to use 30 $^\circ$C and 25 ppt for temperature and salinity was primarily a statistical consideration given a consistent increase over time in the number of days when these thresholds were crossed. That is, sufficient change and variation in the independent variables for the models of seagrass change were needed to statistically describe potential relationships. The reported threshold values in tropical and sub-tropical environments suggest that the limits of the ecological niche for seagrasses are higher for temperature and lower for salinity [@phillips1960; @mcmillan1967; @zieman1975; @lirman2003]. Because we did not see a dramatic increase in the number of days each year when the thresholds were crossed at more stressful values, conditions in Tampa Bay in recent years are likely suboptimal but within the ecological niche for seagrasses. This is especially true for *H. wrightii* that had the greatest changes over the period of record and is tolerant of a wide range of salinity. However, this does not suggest that these factors are unimportant, both currently and in the future. Extreme temperature or precipitation events acting individually or in combination are likely captured by the trends in stressor metrics using these lower thresholds, i.e., an increase in a bay segment median number of days also suggests extremes are increasing given the variation around these summary metrics (@fig-mixeff). Without more continuous, diel observations of these metrics over the period of record, we were heavily reliant on these model outputs to determine relevant thresholds for Tampa Bay seagrass. This further highlights a long-standing data gap and need for the estuary. Regardless, our thresholds seem to be indicative of the potential for chronic sublethal effects of stress on seagrasses, reducing their resilience to other stressors. Further, our models suggested that temperature and salinity changes are at least associated with seagrass loss and, if so, long-term trends in both are set to amplify their effect in the future.

@@

Natural resource managers should consider how these climate-related stressors may alter the effectiveness of intervention activities aimed at protecting ecological resources in Tampa Bay. Management actions that have historically been effective may become less so, resulting in diminished ecosystem resilience compounded by climate change. For example, nitrogen load reductions have historically been the most effective strategy to restore seagrass in Tampa Bay [@greening2006; @greening2014]. As Tampa Bay becomes hotter and fresher, current nutrient load targets may no longer be effective, resulting in further shifts in algal and seagrass ecology dynamics. Strategies that mimic or restore pre-development hydrology or that further reduce allowable load inputs from regulated entities (e.g., additional stormwater controls, hydrological modifications) may be needed to confer additional resilience and adaptive capacity for seagrasses responding to climatic changes in Tampa Bay. These considerations are especially critical for upper Tampa Bay where a majority of seagrass loss has occurred and where temperature and salinity trends appear most pronounced. Reversal of recent trends may be more likely to occur if aggressive actions and controls are pursued sooner rather than later, especially since the challenges of restoring these long-lived foundation species once lost will be exacerbated by ongoing development in the watershed and the current climate trajectory.
